---
title: "Monitoring Early Development of Epidemic Outbreaks with Continuous Time Markov Chains and Bayesian Filtering: Simulations"
author: "MB & RFB"
date: "15/02/2021"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
---

Load Libraries
```{r setup, include = F}
knitr::opts_chunk$set(echo = TRUE)
library(expm)
library(ggplot2)
library(tidyverse)
library(patchwork)
library(Hmisc)
library(sf)
library(scales)
library(knitr)
library(ggtext)
library(readxl)

source("shiny/functions.R")
```

Set color schemes
```{r}
# Fancy colors
branded_colors <- c(
  "blue"   = "#00798c",
  "red"    = "#d1495b",
  "yellow" = "#edae49",
  "green"  = "#66a182",
  "navy"   = "#2e4057", 
  "grey"   = "#8d96a3"
)

AAU_palettes <- list(
  "main" = c("#54616E","#594FBF","#211A52"),
  "second" = c("#bb5b17", "#97701f", "#007fa3", "#a16547", "#0e8563", "#cc445b"),
  "third"  = c("#df8e2e", "#b19336", "#31a9c1", "#cc8b88", "#5caf8d", "#e78293")
)

AAU_pal <- function(palette = "main", reverse = FALSE, ...) {
  pal <- AAU_palettes[[palette]]
  
  if (reverse) pal <- rev(pal)
  
  colorRampPalette(pal, ...)
}
```

Basic statistics for the North Denmark Region, 31 December 2020.
```{r}
north.denmark <- read_xlsx("../Data/Population-North-DenmarkRegion-31122020.xlsx",
                           range = "B4:C14",
                           col_names = c("Municipality", "Inhabitants")) %>% 
  mutate(Type = ifelse(Municipality %in% c("Mariagerfjord", "Morsø","Rebild", "Aalborg"),
                       "non-locked", "Locked")) %>% 
  arrange(Type)

nd.summary <- north.denmark %>% 
  group_by(Type) %>% summarise(Inhabitants = sum(Inhabitants))

N.locked <- as.integer(nd.summary[1,2])
N <- sum(nd.summary["Inhabitants"])
```

# Simulations
## Increased WGS testing activity and restrictions
```{r}
N <- 600000
NumDays = 60
nk      = 10000
pk <- c(0.01, 0.25, 0.5, 0.75)
R <- c(0.5, 0.75, 1.0, 1.25, 1.5)
Threshold = c(0.85,0.90,0.95)

for(j in 1:(length(R))){
  for(i in 1:(length(pk))){
  PP <- runCalc(N1 = N,
                n1 = pk[i]*nk,
                gamma1=0.5,
                R1=R[j],
                MaxI=100,
                NumDays=NumDays,
                TimeIntervention=Inf,
                IniMean=11,
                Y = c(11,rep(0,NumDays-1)),
                Nlow = 0, Nhigh=20,
                IniProb="Uniform")
  tmp <- data.frame(Day=1:NumDays, Prob=PP[1,], Ratio = pk[i], R = R[j])
  if(i==1 && j==1){
    res <- tmp}else 
      res <- rbind(res, tmp)
  }
}
```

Plot results
```{r}
res.testing <- res[res$R==1,] 

res.testing$Ratio <- as.factor(as.character(res.testing$Ratio))
  
p.testing <- ggplot(res.testing, aes(x=Day, y=Prob)) + 
  geom_point(res.testing, mapping = aes(x=Day, y=Prob, color=Ratio)) + 
  xlab("Week") + ylab("Probability of extinction") + 
  geom_hline(yintercept=0.85, linetype="dashed", color = "grey", size = 1) + 
  geom_hline(yintercept=0.9, linetype="dashed", color = "grey", size = 1) +
  geom_hline(yintercept=0.95, linetype="dashed", color = "grey", size = 1) +
  scale_color_manual(values= unname(branded_colors)[c(3,1,2,4)])
p.testing$labels$colour <- "WGS Ratio"

res.restrictions <- res[res$Ratio==0.25,] 

res.restrictions$R <- as.factor(as.character(res.restrictions$R))

p.restrictions <- ggplot(res.restrictions, aes(x=Day, y=Prob)) + 
  geom_point(res.restrictions, mapping = aes(x=Day, y=Prob, color=R)) + 
  xlab("Week") + ylab("Probability of extinction") + 
  geom_hline(yintercept=0.85, linetype="dashed", color = "grey", size = 1) + 
  geom_hline(yintercept=0.9, linetype="dashed", color = "grey", size = 1) +
  geom_hline(yintercept=0.95, linetype="dashed", color = "grey", size = 1) +
  scale_color_manual(values= unname(branded_colors), name = "R<sub>0</sub>") +
  theme(legend.title = element_markdown())
p.restrictions$labels$colour <- "R"

# Figure
p.res <- p.testing / p.restrictions + plot_annotation(tag_levels = "A") 
p.res
```

```{r, include = F}
# Dump figure
ggsave(p.res, filename = "../Output/Figures/Interventions.pdf", width = 18, height = 12, units="cm")
```


Table with results
```{r, warning = F}
thresholds <- res %>% 
  group_by(Ratio, R) %>%
  summarise("Prob < 0.85" = thres(Prob,0.85),
            "Prob < 0.90" = thres(Prob,0.90),
            "Prob < 0.95" = thres(Prob,0.95))

rownames(thresholds) <- NULL
kable(thresholds, caption = "Weeks to thresholds for various testing and restriction strategies.")
```

```{r}
# Dump LaTeX table
col_heads_out <- c("Ratio","R_0", "Prob $<$ 0.85", "Prob $<$ 0.90", "Prob $<$ 0.95")
nrg <- unname(table(thresholds$Ratio))
thresholds2 <- thresholds
thresholds2$Ratio[duplicated(thresholds2$Ratio) == T] <- " "

# storing the table as an object in the workspace prevents compilation, but still saves the .tex
temp_table <- latex(thresholds2,file="../Output/Tables/days.tex", 
                    caption = "Weeks to thresholds for various testing and restriction strategies.", 
                    label = "tab:intervention", rowname=NULL, where = 'tb',
                    colheads = col_heads_out,
                    math.col.names = F,
                    n.rgroup = nrg)
```

## Effect of reproduction rate
```{r}
NumDays = 60
nk      = 10000
pk <- 0.25
R <- seq(0.5, 2.5, by = 0.01)
gamma1 <- 0.5

for(i in 1:length(R)){
  PP <- runCalc(N1 = N,
                n1 = pk*nk,
                gamma1=gamma1,
                R1=R[i],
                MaxI=100,
                NumDays=NumDays,
                TimeIntervention=Inf,
                Y = c(11,rep(0,NumDays-1)),
                Nlow = 0, Nhigh=20,
                IniMean=11,
                IniProb="Uniform")
  tmp <- data.frame(Day=1:NumDays, Prob=PP[1,], Ratio = pk, R = R[i])
  if(i==1){
    res <- tmp}else 
      res <- rbind(res, tmp)
}

thresholds <- res %>% 
  group_by(R) %>%
  summarise("Prob < 0.85" = thres(Prob,0.85),
            "Prob < 0.90" = thres(Prob,0.90),
            "Prob < 0.95" = thres(Prob,0.95))

thresholds.long = thresholds %>%
  gather(`Prob < 0.85`, `Prob < 0.90`, `Prob < 0.95`, key = "Prob", value = "Week")

p.reproduction <- ggplot(thresholds.long,aes(x=R, y=Prob)) + 
  geom_line(thresholds.long, mapping = aes(x=R, y=Week, color = Prob)) +
  xlab("R<sub>0</sub>") + ylab("Week") + 
  scale_color_manual(values = unname(branded_colors)) +
  theme(axis.title.x = element_markdown())
p.reproduction
```

```{r, include= F}
# Dump figure
ggsave(p.reproduction, filename = "../Output/Figures/Reproduction.pdf", width = 18, height = 12, units = "cm")
```


# Danish Mink Case
Map of Denmark
```{r}
geo_sf <- read_sf("https://dawa.aws.dk/kommuner/?format=geojson")
geo_sf <- geo_sf %>% 
  mutate(grp=case_when(navn %in% c("Hjørring","Frederikshavn") ~ "Cluster 5 observed",
                       navn %in% c("Thisted","Brønderslev",
                                   "Vesthimmerlands","Læsø","Jammerbugt") ~ "Locked down",
                       navn %in% c("Morsø", "Rebild","Mariagerfjord","Aalborg") ~ "Not locked down")
  )

fig_dk <- geo_sf %>% ggplot() +
  geom_sf(aes(fill = grp), size=0.1) +
  #scale_fill_manual(values=c(rev(c(AAU_pal("second")(3)[1:3])),"white")) +
  scale_fill_manual(values = c(branded_colors[c("red","blue", "yellow")], "white"), na.translate =F) +
  labs(title="", fill="Status week 46, 2020") +
  theme_void()+
  theme(legend.position=c(0.8,0.8))
fig_dk
```

```{r, include = F}
# Dump figure
ggsave(fig_dk, filename = "../Output/Figures/Denmark.pdf",width = 18, units="cm")
```


## Setup data
Generate Danish mink case
```{r}
# Read Danish Cluster 5 data
c5 <- read.csv("../Data/rn_cluster5_data.csv", sep = ";")
c5$Week <- as.integer(substring(c5$week,7))
c5[is.na(c5)] <- 0
c5 <- c5[c5$Week >= 35 & c5$Week<=53,]
c5.print <- c5

c5.print <- data.frame(Week = c5$Week, PCR = c5$tests, Infected = c5$cases,
                       WGS = c5$sequenced, "Pct" = round(c5$sequenced/c5$cases*100,0), 
                       Cluster5 = c5$cluster5)
rownames(c5.print) <- NULL

```

Table with results
```{r, warning = F}
kable(c5.print, caption = "Weekly test data.")
```

```{r}
# Dump LaTeX table
# storing the table as an object in the workspace prevents compilation, but still saves the .tex
temp_table <- latex(c5.print,file="../Output/Tables/summary.tex", 
                    caption = "Weekly test data.", 
                    label = "tab:summary", rowname=NULL,where='tb')
```

Create intervention table
```{r}
intervention <- data.frame(Week = c(46,47,50),
                           Announced = as.Date(c("11/06/20","11/13/20","11/06/20"), "%m/%d/%y"),
                           Effect = as.Date(c("11/09/20","11/16/20","12/07/20"), "%m/%d/%y"),
                           Description = c("Lockdown", "Reopening", "Planned"))
```

## Analyse data
It was the plan to test the entire population of the region over a 4 weeks period
i.e., nk = N.locked/4 = 71510 PCR tests per week. Assume a postive pct. of 1.5%, i.e,
1072 positive tests. The test capacity was up to 5000 a week, so we assume all 1072
will be WGS sequence, i.e. n2=1072.
```{r}
R1 <- 1.2                   # Reproduction rate before intervention
R2 <- 1                   # Reproduction rate after intervention
gamma1 <- 1.0               # Average contamination time 2 weeks
```

Planned: If restrictions and testing have been realized
```{r}
PP <- runCalc(N1 = N.locked,
        n1 = round(c5$cases[1:11]/2),
        n2 =  1072,
        gamma1=gamma1,
        R1=R1,
        R2=R2,
        MaxI=100,
        NumDays=dim(c5)[1],
        Y = c5$cluster5,
        TimeIntervention=12,
        IniMean=4,
        Nlow = 0, Nhigh=20,
        IniProb="Uniform")

res.mink <- data.frame(Week=c5$Week, Prob=PP[1,], 
                       Strategy="Planned")
```

Realized: Actual observed data. For the realized scenario we also assume a change in the reproduction rate, due to the travel restrictions, but we instead use the actual number of sequenced individuals.
```{r}
PP <- runCalc(N1 = N.locked,
              n1 = round(c5$cases[1:11]/2),
              n2 = round(c5$cases[12:nrow(c5)]/2),
              gamma1=gamma1,
              R1=R1,
              R2=R2,
              MaxI=100,
              NumDays=dim(c5)[1],
              Y = c5$cluster5,
              TimeIntervention=12,
              IniMean=4,
              Nlow = 0, Nhigh=20,
              IniProb="Uniform")


res.mink <- rbind(res.mink, data.frame(Week=c5$Week, Prob=PP[1,], 
                          Strategy="Realized"))

res.mink$Strategy <- factor(res.mink$Strategy)

p.minkcase <- ggplot(res.mink, aes(x=Week, y=Prob, color=Strategy)) + 
  geom_point(size=2) +
  geom_vline(xintercept=45, linetype="dashed", color = "grey", size = 1) +
  geom_vline(xintercept=47, linetype="dashed", color = "grey", size = 1) + 
  geom_vline(xintercept=50, linetype="dashed", color = "grey", size = 1) + 
  xlab("Week") + ylab("Probability of extinction") +
  scale_color_manual(values= unname(branded_colors))
p.minkcase
```

```{r, include = F}
# dump figure
ggsave(p.minkcase, filename = "../Output/Figures/Mink.pdf", width = 18, height = 12, units="cm")
```

Table of Cuts
```{r}
cut.week <- function(x,week){
  x[x$Week == week,]
}

cut.mink <- rbind(cut.week(res.mink,45),
      cut.week(res.mink,47),
      cut.week(res.mink,50))

cut.mink$Prob <- round(cut.mink$Prob,4)

cut.mink.wide <- cut.mink %>% spread(Strategy, Prob)
kable(cut.mink.wide, caption = "Planned and realized probability of extinction, before, at the break, and
      at time of planned intervention")
```

```{r, include = F}
# Dump LaTeX table
temp_table2 <- latex(cut.mink.wide,file="../Output/Tables/cut.tex", 
                     caption = "Planned and realized probability of extinction, before, at the break, and
      at time of planned intervention",  label = "tab:min", rowname=NULL, where = 'tb')

```

```{r}
cut.mink.wide <- round(cut.mink.wide,2)

# Combine figure and table
p.minkcase2 <- p.minkcase +
  annotate(geom = "text", label = cut.mink.wide[1,2], x = cut.mink.wide[1,1] -1.1,
           y = cut.mink.wide[1,2], colour = branded_colors[1]) +
  annotate(geom = "text", label = cut.mink.wide[1,3], x = cut.mink.wide[1,1] +1.1,
           y = cut.mink.wide[1,3], colour = branded_colors[2]) +
  annotate(geom = "text", label = cut.mink.wide[2,2], x = cut.mink.wide[2,1] -1.1,
           y = cut.mink.wide[2,2], colour = branded_colors[1]) +
  annotate(geom = "text", label = cut.mink.wide[2,3], x = cut.mink.wide[2,1] +1.1,
           y = cut.mink.wide[2,3], colour = branded_colors[2]) +
  annotate(geom = "text", label = cut.mink.wide[3,2], x = cut.mink.wide[3,1] -1.1,
           y = cut.mink.wide[3,2], colour = branded_colors[1]) +
  annotate(geom = "text", label = cut.mink.wide[3,3], x = cut.mink.wide[3,1] +1.1,
           y = cut.mink.wide[3,3], colour = branded_colors[2])

p.minkcase2
ggsave(p.minkcase2, filename = "../Output/Figures/Mink2.pdf", width = 18, height = 12, units="cm")
```

## Kullback-Leibler
Simulation of the sampling distribution and approximations
```{r, cache = T}
#N   <- 589148                 # Population size (North Denmark Region)
xkmm <- 0.006*N               # 0.6% of the population is positive (median)
xk  <- round(0.006*N*0.08)    # 8% of infected with specific variant
xkm <- xkmm - xk              # Number of infected with non-specific variant
pk <- 0.28                    # Ratio of positives WGS tested

nk  <- 17000   # Numer of PCR tests (median)
mk  <- round(pk*nk*(xk+xkm)/N)  # 0.28 of positive sent for WGS tests (median)


# Simulation
nsim <- 10000
yk <- cluster.sampling(N, xk, xkm, nk, mk, nsim = nsim)
dyk <- as.data.frame(table(yk))
dyk$yk <- as.integer(levels(dyk$yk))
dyk$Freq <- dyk$Freq/nsim
dyk$Type = "Sample"

# Poisson approximation
lam <- mk*xk/(xk+xkm)
dzkpois <- data.frame(yk=dyk$yk, Freq=dpois(dyk[,1],lam), Type = "Poisson")
KLpois <- kl(dyk, dzkpois)

# Binomial approximation
dzk.binom1 <- data.frame(yk=dyk$yk, Freq =dbinom(dyk[,1], mk, xk/(xk+xkm)), Type = "Binom1")
KLbinom1 <- kl(dyk, dzk.binom1)

# Alternative binomial approximation
dzk.binom2 <- data.frame(yk=dyk$yk, Freq =dbinom(dyk[,1], pk*nk, xk/N), Type = "Binom2")
KLbinom2 <- kl(dyk, dzk.binom2)

# Gather data
dyk.all <- rbind(dyk, dzkpois, dzk.binom1, dzk.binom2)
```

Kullback-Leibler divergence between sampling distribution and plot
```{r}
p.distr <- ggplot(dyk.all, aes(x=yk,y=Freq)) +
  geom_point(dyk.all, mapping = aes(x=yk,y=Freq,color=Type), size = 2) +
  scale_color_manual(values=unname(branded_colors)) +
  xlab("Number") + ylab("Probability") +
  scale_x_continuous(breaks = 0:9)
p.distr
```

```{r, include = F}
# Dump figure
ggsave(p.distr, filename = "../Output/Figures/Distribution.pdf", width = 18, height = 12, units="cm")
```


```{r}
#Try with overlay of KL
p.distr2 <- ggplot(dyk.all, aes(x=yk,y=Freq)) +
  geom_point(dyk.all, mapping = aes(x=yk,y=Freq,color=Type), size = 2) +
  scale_color_manual(values=unname(branded_colors)) +
  xlab("Number") + ylab("Probability") +
  scale_x_continuous(breaks = 0:9) +
  annotate(geom = "Text", x = 7, y = 0.25, label = paste("KL divergence: \n",
                                        "Binom1 =", round(KLbinom1,4), "\n",
                                        "Binom2 =", round(KLbinom2,4), "\n",
                                        "Poisson =", round(KLpois,4)),
           hjust = 0)
p.distr2
ggsave(p.distr2, filename = "../Output/Figures/Distribution2.pdf", width = 18, height = 12, units="cm")
```


Table
```{r}
kld <- data.frame(Type = c("Poisson", "Binom1", "Binom2"),
                 KLD  = round(c(KLpois, KLbinom1, KLbinom2),4))

#kld <- dyk.all %>% pivot_wider(names_from = "Type", values_from = "Freq") %>% 
#  rename("Number" = yk)

kable(kld, caption = "Kulback-Leibler divergence between the approximations and the simulated sampling distribution. Parameters used were $N=600,000$, $x_k=3,312$, $x_k'=288$, $n_k=17,000$, and $m_k=29$.")
```


```{r, include = F}
# Dump latex table
temp_table <- latex(kld,file="../Output/Tables/kld.tex", 
                    caption = "Kulback-Leibler divergence between the approximations and the simulated sampling distribution. Parameters used were $N=600,000$, $x_k=3,312$, $x_k'=288$, $n_k=17,000$, and $m_k=29$.",
                    label = "tab:kld", rowname=NULL, where='tb')
```


```{r}

prior <- function(a, alpha = 1, p = 10^-3){
  1/(1+exp(a/alpha))
}

extiction <- function( a, alpha = 1, N = 15, p = 10^-3){
  1/(1+exp(a/alpha)*(1-p)^exp(a*N))
}

pr <- -100:20/10
ap <- -1000:200/100

extinction <- data.frame(a=ap, prob = extiction(ap), Probability = "A-posteriori")
prior <- data.frame(a=pr, prob=prior(pr), Probability = "A-priori")

prob <- rbind(extinction, prior)
prob$Probability <- factor(prob$Probability)

p.extinction <- ggplot(prob, aes(x=a, y=prob, group = Probability, col = Probability))+
  geom_line(data=prob[prob$Probability == "A-posteriori", ]) +
  geom_point(data=prob[prob$Probability == "A-priori", ]) +
    scale_color_manual(values = unname(branded_colors)) +
  xlab("Growth rate") + 
  ylab("Probability of extinction")

p.extinction

ggsave(p.extinction, filename = "../Output/Figures/Extinction.pdf", width = 18, height = 12, units="cm")

```



