---
title: "Monitoring Early Development of Epidemic Outbreaks with Continuous Time Markov Chains and Bayesian Filtering: Simulations"
author: "MB & RFB"
date: "15/02/2021"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
---

Load Libraries
```{r setup, include = F}
knitr::opts_chunk$set(echo = TRUE)
library(expm)
library(ggplot2)
library(tidyverse)
library(patchwork)
library(Hmisc)
library(sf)
library(scales)
library(knitr)

source("shiny/functions.R")
```

Set color schemes
```{r}
# Fancy colors
branded_colors <- c(
  "blue"   = "#00798c",
  "red"    = "#d1495b",
  "yellow" = "#edae49",
  "green"  = "#66a182",
  "navy"   = "#2e4057", 
  "grey"   = "#8d96a3"
)

AAU_palettes <- list(
  "main" = c("#54616E","#594FBF","#211A52"),
  "second" = c("#bb5b17", "#97701f", "#007fa3", "#a16547", "#0e8563", "#cc445b"),
  "third"  = c("#df8e2e", "#b19336", "#31a9c1", "#cc8b88", "#5caf8d", "#e78293")
)

AAU_pal <- function(palette = "main", reverse = FALSE, ...) {
  pal <- AAU_palettes[[palette]]
  
  if (reverse) pal <- rev(pal)
  
  colorRampPalette(pal, ...)
}
```

Basic statistics for the North Denmark Region, 31 December 2020.
```{r}
north.denmark <- data.frame(Municipality = c("Brønderslev", "Frederikshavn", "Hjørring",
                                             "Jammerbugt","Læsø", "Vesthimmerlands", 
                                             "Thisted", "Mariagerfjord", "Morsø",
                                             "Rebild", "Aalborg"),
                            Inhabitants  = c(36304, 59654, 64483, 38324, 1786, 36727, 43423, 41800, 20247, 
                                             30113, 217075),
                            Type = c(rep("Locked",7),rep("non-Locked",4)))

nd.summary <- north.denmark %>% 
  group_by(Type) %>% summarise(Inhabitants = sum(Inhabitants))

N.locked <- as.integer(nd.summary[1,2])
N <- sum(nd.summary["Inhabitants"])
```

# Simulations
## Increased WGS testing activity and restrictions
```{r}
N <- 600000
NumDays = 60
nk      = 10000
pk <- c(0.25, 0.5, 0.75, 1.0)
R <- c(0.5, 0.75, 1.0, 1.25, 1.5)
Threshold = c(0.85,0.90,0.95)

for(j in 1:(length(R))){
  for(i in 1:(length(pk))){
  PP <- runCalc(N1 = N,
                n1 = pk[i]*nk,
                gamma1=0.5,
                R1=R[j],
                MaxI=100,
                NumDays=NumDays,
                TimeIntervention=Inf,
                IniMean=11,
                IniProb="Kronecker")
  tmp <- data.frame(Day=1:NumDays, Prob=PP[1,], Ratio = pk[i], R = R[j])
  if(i==1 && j==1){
    res <- tmp}else 
      res <- rbind(res, tmp)
  }
}
```

Plot results
```{r}
res.testing <- res[res$R==1,] 

res.testing$Ratio <- as.factor(as.character(res.testing$Ratio))
  
p.testing <- ggplot(res.testing, aes(x=Day, y=Prob)) + 
  geom_point(res.testing, mapping = aes(x=Day, y=Prob, color=Ratio)) + 
  xlab("Week") + ylab("Probability of extinction") + 
  geom_hline(yintercept=0.85, linetype="dashed", color = "grey", size = 1) + 
  geom_hline(yintercept=0.9, linetype="dashed", color = "grey", size = 1) +
  geom_hline(yintercept=0.95, linetype="dashed", color = "grey", size = 1) +
  scale_color_manual(values= unname(branded_colors))
p.testing$labels$colour <- "WGS Ratio"

res.restrictions <- res[res$Ratio==0.25,] 

res.restrictions$R <- as.factor(as.character(res.restrictions$R))

p.restrictions <- ggplot(res.restrictions, aes(x=Day, y=Prob)) + 
  geom_point(res.restrictions, mapping = aes(x=Day, y=Prob, color=R)) + 
  xlab("Week") + ylab("Probability of extinction") + 
  geom_hline(yintercept=0.85, linetype="dashed", color = "grey", size = 1) + 
  geom_hline(yintercept=0.9, linetype="dashed", color = "grey", size = 1) +
  geom_hline(yintercept=0.95, linetype="dashed", color = "grey", size = 1) +
  scale_color_manual(values= unname(branded_colors))
p.restrictions$labels$colour <- "R"

# Figure
p.res <- p.testing / p.restrictions + plot_annotation(tag_levels = "A") 
p.res
```

```{r, include = F}
# Dump figure
ggsave(p.res, filename = "../Output/Figures/Interventions.pdf", width = 18, units="cm")
```


Table with results
```{r, warning = F}
thresholds <- res %>% 
  group_by(Ratio, R) %>%
  summarise("Prob < 0.85" = thres(Prob,0.85),
            "Prob < 0.90" = thres(Prob,0.90),
            "Prob < 0.95" = thres(Prob,0.95))

rownames(thresholds) <- NULL
kable(thresholds, caption = "Weeks to thresholds for various testing and restriction strategies.")
```

```{r}
# Dump LaTeX table
# storing the table as an object in the workspace prevents compilation, but still saves the .tex
temp_table <- latex(thresholds,file="../Output/Tables/days.tex", 
                    caption = "Weeks to thresholds for various testing and restriction strategies.", 
                    label = "tab:intervention", rowname=NULL, where = 'tb')
```

## Effect of reproduction rate
```{r}
NumDays = 60
nk      = 10000
pk <- 0.25
R <- seq(0.5, 2.5, by = 0.01)
gamma1 <- 0.5

for(i in 1:length(R)){
  PP <- runCalc(N1 = N,
                n1 = pk*nk,
                gamma1=gamma1,
                R1=R[i],
                MaxI=100,
                NumDays=NumDays,
                TimeIntervention=Inf,
                IniMean=12,
                IniProb="Kronecker")
  tmp <- data.frame(Day=1:NumDays, Prob=PP[1,], Ratio = pk, R = R[i])
  if(i==1){
    res <- tmp}else 
      res <- rbind(res, tmp)
}

thresholds <- res %>% 
  group_by(R) %>%
  summarise("Prob < 0.85" = thres(Prob,0.85),
            "Prob < 0.90" = thres(Prob,0.90),
            "Prob < 0.95" = thres(Prob,0.95))

thresholds.long = thresholds %>%
  gather(`Prob < 0.85`, `Prob < 0.90`, `Prob < 0.95`, key = "Prob", value = "Week")

p.reproduction <- ggplot(thresholds.long,aes(x=R, y=Prob)) + 
  geom_line(thresholds.long, mapping = aes(x=R, y=Week, color = Prob)) +
  xlab("R") + ylab("Week") + 
  scale_color_manual(values = unname(branded_colors))
p.reproduction
```

```{r, include= F}
# Dump figure
ggsave(p.reproduction, filename = "../Output/Figures/Reproduction.pdf", width = 18, units = "cm")
```


# Danish Mink Case
Map of Denmark
```{r}
geo_sf <- read_sf("https://dawa.aws.dk/kommuner/?format=geojson")
geo_sf <- geo_sf %>% 
  mutate(grp=case_when(navn %in% c("Hjørring","Frederikshavn") ~ "Cluster 5",
                       navn %in% c("Thisted","Brønderslev",
                                   "Vesthimmerlands","Læsø","Jammerbugt") ~ "Locked down",
                       navn %in% c("Morsø", "Rebild","Mariagerfjord","Aalborg") ~ "North Denmark",
                       TRUE ~ "Remaining"))

fig_dk <- geo_sf %>% ggplot() +
  geom_sf(aes(fill = grp), size=0.1) +
  #scale_fill_manual(values=c(rev(c(AAU_pal("second")(3)[1:3])),"white")) +
  scale_fill_manual(values = c(branded_colors[c("red","blue", "yellow")], "white")) +
  labs(title="", fill="Status week 46, 2020") +
  theme_void()+
  theme(legend.position=c(0.8,0.8))
fig_dk
```

```{r, include = F}
# Dump figure
ggsave(fig_dk, filename = "../Output/Figures/Denmark.pdf",width = 18, units="cm")
```


## Setup data
Generate Danish mink case
```{r}
# Read Danish Cluster 5 data
c5 <- read.csv("../Data/rn_cluster5_data.csv", sep = ";")
c5$Week <- as.integer(substring(c5$week,7))
c5[is.na(c5)] <- 0
c5 <- c5[c5$Week >= 35 & c5$Week<=53,]
c5.print <- c5

c5.print <- data.frame(Week = c5$Week, PCR = c5$tests, Infected = c5$cases,
                       WGS = c5$sequenced, Cluster5 = c5$cluster5)
rownames(c5.print) <- NULL

```

Table with results
```{r, warning = F}
kable(c5.print, caption = "Weekly test data.")
```

```{r}
# Dump LaTeX table
# storing the table as an object in the workspace prevents compilation, but still saves the .tex
temp_table <- latex(c5.print,file="../Output/Tables/summary.tex", 
                    caption = "Weekly test data.", 
                    label = "tab:summary", rowname=NULL,where='tb')
```

Create intervention table
```{r}
intervention <- data.frame(Week = c(46,47,50),
                           Announced = as.Date(c("11/06/20","11/13/20","11/06/20"), "%m/%d/%y"),
                           Effect = as.Date(c("11/09/20","11/16/20","12/07/20"), "%m/%d/%y"),
                           Description = c("Lockdown", "Reopening", "Planned"))
```

## Analyse data
It was the plan to test the entire population of the region over a 4 weeks period
i.e., nk = N.locked/4 = 71510 PCR tests per week. Assume a postive pct. of 1.5%, i.e,
1072 positive tests. The test capacity was up to 5000 a week, so we assume all 1072
will be WGS sequence, i.e. n2=1072
```{r}
R1 <- 1.2                   # Reproduction rate before intervention
R2 <- 1                   # Reproduction rate after intervention
gamma1 <- 1.0               # Average contamination time 2 weeks
```

Planned: If restrictions and testing have been realized
```{r}
PP <- runCalc(N1 = N.locked,
        n1 = round(c5$cases[1:11]/2),
        n2 =  1072,
        gamma1=gamma1,
        R1=R1,
        R2=R2,
        MaxI=100,
        NumDays=dim(c5)[1],
        Y = c5$cluster5,
        TimeIntervention=12,
        IniMean=4,
        IniProb="Kronecker")

res.mink <- data.frame(Week=c5$Week, Prob=PP[1,], 
                       Strategy="Planned")
```

Realized: Actual observed data
```{r}

PP <- runCalc(N1 = N.locked,
              n1 = round(c5$cases/2),
              gamma1=gamma1,
              R1=R1,
              R2=R2,
              MaxI=100,
              NumDays=dim(c5)[1],
              Y = c5$cluster5,
              TimeIntervention=Inf,
              IniMean=4,
              IniProb="Kronecker")

res.mink <- rbind(res.mink, data.frame(Week=c5$Week, Prob=PP[1,], 
                          Strategy="Realized"))

res.mink$Strategy <- factor(res.mink$Strategy)

p.minkcase <- ggplot(res.mink, aes(x=Week, y=Prob)) + 
  geom_point(res.mink, mapping = aes(x=Week, y=Prob, color=Strategy), size=2)+
  geom_vline(xintercept=45, linetype="dashed", color = "grey", size = 1) +
  geom_vline(xintercept=47, linetype="dashed", color = "grey", size = 1) + 
  geom_vline(xintercept=50, linetype="dashed", color = "grey", size = 1) + 
  xlab("Week") + ylab("Probability of extinction") +
  scale_color_manual(values= unname(branded_colors))
p.minkcase
```

```{r, include = F}
# dump figure
ggsave(p.minkcase, filename = "../Output/Figures/Mink.pdf", width = 18, units="cm")
```

Table of Cuts
```{r}
cut.week <- function(x,week){
  x[x$Week == week,]
}

cut.mink <- rbind(cut.week(res.mink,45),
      cut.week(res.mink,47),
      cut.week(res.mink,50))

cut.mink$Prob <- round(cut.mink$Prob,4)

cut.mink.wide <- cut.mink %>% spread(Strategy, Prob)
kable(cut.mink.wide, caption = "Planned and realized probability of extinction, before, at the break, and
      at time of planned intervention")
```


```{r, include = F}
# Dump LaTeX table
temp_table2 <- latex(cut.mink.wide,file="../Output/Tables/cut.tex", 
                     caption = "Planned and realized probability of extinction, before, at the break, and
      at time of planned intervention",  label = "tab:min", rowname=NULL, where = 'tb')

```

## Kullback-Leibler
Simulation of the sampling distribution and approximations
```{r, cache = T}
#N   <- 589148                 # Population size (North Denmark Region)
xkmm <- 0.006*N               # 0.6% of the population is positive (median)
xk  <- round(0.006*N*0.08)    # 8% of infected with specific variant
xkm <- xkmm - xk              # Number of infected with non-specific variant
pk <- 0.28                    # Ratio of positives WGS tested

nk  <- 17000   # Numer of PCR tests (median)
mk  <- round(pk*nk*(xk+xkm)/N)  # 0.28 of positive sent for WGS tests (median)


# Simulation
nsim <- 10000
yk <- cluster.sampling(N, xk, xkm, nk, mk, nsim = nsim)
dyk <- as.data.frame(table(yk))
dyk$yk <- as.integer(levels(dyk$yk))
dyk$Freq <- dyk$Freq/nsim
dyk$Type = "Sample"

# Poisson approximation
lam <- mk*xk/(xk+xkm)
dzkpois <- data.frame(yk=dyk$yk, Freq=dpois(dyk[,1],lam), Type = "Poisson")
KLpois <- kl(dyk, dzkpois)

# Binomial approximation
dzk.binom1 <- data.frame(yk=dyk$yk, Freq =dbinom(dyk[,1], mk, xk/(xk+xkm)), Type = "Binom1")
KLbinom1 <- kl(dyk, dzk.binom1)

# Alternative binomial approximation
dzk.binom2 <- data.frame(yk=dyk$yk, Freq =dbinom(dyk[,1], pk*nk, xk/N), Type = "Binom2")
KLbinom2 <- kl(dyk, dzk.binom2)

# Gather data
dyk.all <- rbind(dyk, dzkpois, dzk.binom1, dzk.binom2)
```

Kullback-Leibler divergence between sampling distribution and plot
```{r}
p.distr <- ggplot(dyk.all, aes(x=yk,y=Freq)) +
  geom_point(dyk.all, mapping = aes(x=yk,y=Freq,color=Type), size = 2) +
  scale_color_manual(values=unname(branded_colors)) +
  xlab("Number") + ylab("Probability")
p.distr
```

```{r, include = F}
# Dump figure
ggsave(p.distr, filename = "../Output/Figures/Distribution.pdf", width = 18, units="cm")
```


Table
```{r}
kld <- data.frame(Type = c("Poisson", "Binom1", "Binom2"),
                 KLD  = round(c(KLpois, KLbinom1, KLbinom2),4))
kable(kld, caption = "Kulback-Leibler divergence between the approximations and the simulated sampling distribution")
```


```{r, include = F}
# Dump latex table
temp_table <- latex(kld,file="../Output/Tables/kld.tex", 
                    caption = "Kulback-Leibler divergence between the approximations and the simulated sampling distribution.",
                    label = "tab:kld", rowname=NULL, where='tb')
```

